---
title: "Cladosporium experiment 2"
author: "Devin R. Leopold"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

Load packages and data
```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(ggtext)
library(brms)
library(bayestestR)
library(ggdist)
library(ggpubr)

#Import raw data
dat <- read.csv("data/Experiment2.csv")
dat$treatment[dat$treatment=="Pos. control"] <- "Live"

#Subset experimental treatment samples
dat.expt <- dat %>% filter(!is.na(isolate)) %>%
  mutate(plate=paste(plate,complex,sep="-"))

#Subset negative control samples
dat.neg <- dat %>% filter(is.na(isolate)) %>%
  mutate(plate=paste(plate,complex,sep="-"))
```

# Fit negative binomial mixed-model with brms.
Fixed effects in this model include the experimental treatment, the species complex of the *Cladosporium* isolates, and their interaction. Isolate identity and plate (each holding 2-3 leaf disks) and included as random intercept terms. 

```{r fig.height=10, fig.width=8, warning=F, message=F, results='hide', cache=T}
model_e2 <- brm(
  disease_severity ~ treatment * complex + (1|plate) + (1|isolate),
  data = dat.expt, 
  family = negbinomial(), 
  chains = 4, cores = 4, iter = 4000,
  control = list(adapt_delta = 0.999),
  prior=prior(cauchy(0, 2.5), class='b'),
  sample_prior = T,
  save_pars = save_pars(all = TRUE),
  silent = 2, refresh = 0,
  seed = 80
)
plot(model_e2, N=9)
```

We can validate the model by comparing simulated data from the model to the observed data (ie posterior predictive checks).

```{r}
pp_check(model_e2, nsamples = 100)
```

We can also check for overly influential data points using Pareto-smoothed importance sampling (PSIS).

```{r, warning=FALSE}
(loo_rust <- loo(model_e2, cores=parallel::detectCores(), moment_match = TRUE))
plot(loo_rust)
```

# Fit negative control model
Because negative control samples do not involve any isolates, they do not fit the main model structure. In order to use these samples as a reference point we can instead fit a simple, intercept only model (with a plate-level random intercept), with an negative binomial error structure in brms. We can then use the posterior distribution and its HDI as a reference against which we can compare the other treaments

```{r fig.height=5, fig.width=8, warning=F, message=F, results='hide', cache=T}
model_neg <- brm(
  disease_severity ~ 1 + (1|plate),
  data = dat.neg, 
  family = negbinomial(), 
  chains = 4, cores = 4, iter = 4000,
  control = list(adapt_delta = 0.999),
  silent = 2, refresh = 0,
  seed = 80
)
plot(model_neg, N=3)
```

We can validate the model by comparing simulated data from the model to the observed data (ie posterior predictive checks).

```{r}
pp_check(model_neg, nsamples = 100)
```

****

# Results
We will visualize the results by showing the posterior distribution and 89% HDIs for each treatment and species complex combination. Within each species complex we will present the probability of direction and the 89% HDI of the decrease in disease severity proceeding from Autoclaved -> Filtered -> Live *Cladosporium* inoculum. We will also overlay the median and 89% HDI of disease severity for the negative control (sterile water) samples.


```{r fig.width=7, fig.height=4.5, echo=F}
# Extract posterior distribution of negative control samples
neg_pr <- posterior_epred(model_neg, re_formula=NA, resp="Intercept")

# Posterior distributions of experimental treatments x species complex
d_pred <- dat.expt %>%
  dplyr::select(complex, treatment) %>%
  distinct() %>%
  mutate(combo=paste(complex,treatment,sep="_"))
pr <- posterior_epred(model_e2, newdata=d_pred, re_formula=NA)
colnames(pr) <- d_pred$combo

# Make the figure
p <- pr %>% 
  data.frame %>% pivot_longer(everything()) %>%
  separate(name, into=c("complex", "treatment"), sep="_") %>%
  mutate(
    complex=case_when(
      complex=='herbarum' ~ "*C. herbarum* complex",
      complex=='cladosporioides' ~ "*C. cladosporioides* complex"
  )) %>%
  ggplot(aes(x=treatment, y=value)) +
  stat_histinterval(slab_type = 'pdf', trim = T, scale=0.5,
                    .width = c(0.89), point_interval='median_hdi') +
  geom_hline(yintercept = hdi(neg_pr[,1], .width=0.89)[1], linetype='dotted') +
  geom_hline(yintercept = hdi(neg_pr[,1], .width=0.89)[2], linetype='dotted') +
  geom_hline(yintercept = median(neg_pr[,1])) +
  #annotate("text", x = 1, y = 20, label = "Arbitrary text") +
  scale_y_continuous(breaks=c(25, 50, 75, 100, 125)) +
  coord_cartesian(ylim=c(5, 160), clip='off') +
  facet_wrap(~complex) +
  labs(y="Disease severity (# of uredinia)") +
  ggthemes::theme_few() +
  theme(legend.position = 'none',
        axis.title.x = element_blank(),
        axis.text = element_text(color="black"),
        strip.background = element_blank(),
        strip.text = element_markdown(size=13, hjust = 0))

# Define a function to calculate treatment comparisons and add to the figure
compareTreatments <- function(gg, y, a, b, complex, pr){
  pdist <- pr %>% data.frame %>%
    transmute(pdist=!!sym(a)-!!sym(b)) %$% pdist
  PD <- pd(pdist) %>% as.numeric %>%
    multiply_by(100) %>% round(1)
  HDI <- hdi(pdist, .width=0.89) %>% 
    apply(1:2, function(x)round(x,1))
  df <- data.frame(
    y=y, 
    complex=complex,
    group1=str_extract(a, "[^_]+$"),
    group2=str_extract(b, "[^_]+$"),
    PD=PD,
    HDI=paste(HDI,collapse = "-")
  )
  gg + stat_pvalue_manual(df, label="{PD}% pd\n[{HDI}]", y.position = "y", size=3.25,
                          tip.length = 0.01, bracket.shorten = 0.1, vjust=-0.1)
}

# Add summary stats of between treatment comparisons
p %<>% compareTreatments(115, "cladosporioides_Autoclaved","cladosporioides_Filtered",
               "*C. cladosporioides* complex", pr) %>%
  compareTreatments(115, "herbarum_Autoclaved","herbarum_Filtered",
               "*C. herbarum* complex", pr) %>%
  compareTreatments(140, "cladosporioides_Autoclaved","cladosporioides_Live",
               "*C. cladosporioides* complex", pr) %>%
  compareTreatments(140, "herbarum_Autoclaved","herbarum_Live",
               "*C. herbarum* complex", pr) %>%
  compareTreatments(90, "cladosporioides_Filtered","cladosporioides_Live",
               "*C. cladosporioides* complex", pr) %>%
  compareTreatments(90, "herbarum_Filtered","herbarum_Live",
               "*C. herbarum* complex", pr)

# Compare treatments with negative control
neg_pr <- pr %>%
  data.frame %>%
  mutate(neg=neg_pr[,1]) %>%
  mutate_at(vars(-matches("neg")), list(~ neg - .)) %>%
  select(-neg) 

neg_DF <- data.frame(
  treatment=colnames(neg_pr),
  pd = apply(neg_pr, 2, function(x) bayestestR::pd(x) ) %>% unname %>%
    multiply_by(100) %>% round(1),
  HDI = apply(neg_pr, 2, function(x) paste(round(as.numeric(bayestestR::hdi(x, ci=0.89)[2:3]),1), sep="",collapse = "-") ),
  row.names = NULL) 

negStats <- data.frame(
  complex=rep(c("*C. herbarum* complex","*C. cladosporioides* complex"), each=3),
  treatment=rep(c("Autoclaved","Filtered","Live"),2),
  lab=c(
    paste0(neg_DF$pd[1], "% pd\n[",neg_DF$HDI[1],"]"),
    paste0(neg_DF$pd[2], "% pd\n[",neg_DF$HDI[2],"]"),
    paste0(neg_DF$pd[3], "% pd\n[",neg_DF$HDI[3],"]"),
    paste0(neg_DF$pd[4], "% pd\n[",neg_DF$HDI[4],"]"),
    paste0(neg_DF$pd[5], "% pd\n[",neg_DF$HDI[5],"]"),
    paste0(neg_DF$pd[6], "% pd\n[",neg_DF$HDI[6],"]")
  )
)

p <- p + geom_text(data=negStats, aes(y=10, label=lab), size=3.25) 

# Save and print
ggsave( "output/figure_expt2.jpg", width=7, height=4.5)
p
```

